---
title: "Experiment 3: More Airline Data"
subtitle: "Asynchronous Lecture"
date: "2023-09-10"
output: html_notebook
---
Time Stamp: 32:00
Alexis Alexander

This experiment (concerptually) continues our class work on flight data. 

We will continue to refer to (https://r4ds.hadley.nz/data-transform).

**However, we will expand and update our dataset by going to a root source -- the Bureau of Transportation.

Go to the website https://www.transtats.bts.gov/DL_SelectFields.aspx?gnoyr_VQ=FIM&QO_fu146_anzr=Nv4%20Pn44vr45. This dataset allows access to information on domestic flight and domestic carriers, with monthly summary data. 

# Packages


```{r}
library(nycflights13)
library(tidyverse)
library(janitor)
library(here)
```

# Get data from the online source.

Save the download to your data_raw directory. It will be a zipfile, but that will be OK.
Don't change the name on the zipfile.

## Load the BOT data 

We take advantage of the here package to get the path to the file that we stored. We also may not that inside the zipfile is a csv file. We will read the csv file.

```{r}
thisfile=here("data_raw", "DL_SelectFields.zip")

df2022=read_csv(thisfile) %>% clean_names()
```
## Subsetting to data of interest 

Let's focus on flights from LGA and eliminate cargo flights by requiring at least 1 passenger. 

Call the resultant dataframe 'df9'.

```{r}
df9=df2022 |> filter(passengers>0,origin=="LGA")
```

 # Exploring the BOT 2022 data
 
 ## Subsetting to data of interest 
 
 Lets try to visualize a monthly count of flights using a barchart.
 
 First, a not quite right visualization 
 
 aes - declare what is on the horizontal, and geometry is bar 
 
```{r}
df9 |>
  ggplot(aes(month)) + geom_bar()
```
 This shows flights from everywhere out of LGA. The lines in tables are counting the number of departures in a month. Count the number of departures and not the number of rows. 
 
 By default, geom_bar is countign the number of rows, where we have asked it to visualize the count by moth 
 
 ** Take a look at the dattset and discover why counting rows is not going to give us a count of flights***
 
 weight each row by a certain value, weigh each  row by the actual number of departures 
  
The visualization we want is: 

```{r}
df9 |> ggplot(aes(month)) + geom_bar(aes(weight=departures_performed))
```
Make some observations about this plot. 

### A New Visualization 

Can you make a boxplot where the bar height is based on the number of passengers riding each month?

```{r}
df9|> ggplot(aes(month)) + geom_bar(aes(weight=passengers))
```

Observation:
* Low passenger numbers in early 2022 likely due to Covid
* Flights may still have been affected by the pandemic 


***Just Because You Can 

```{r}
df9 |>
  ggplot(aes(month, fill=carrier_name)) + geom_bar(aes(weight=departures_performed))
```

## Arrivals and Departures from LGA

```{r}
df10=df2022 |> filter(passengers>0,origin=="LGA" | dest=="LGA")

df10 |> ggplot(aes(month)) + geom_bar(aes(weight=passengers))
```

originally selecting by name but could select based on a subset of these variables
```{r}
df11=df10 |>
  select(month,passengers,seats,carrier_name,dest,origin)
```

```{r}
df12=df10 |> select(1:5, month, contains("id"))
```

How full was the airplane?
Distribution of how full these airlines were
```{r}
df13=df11 |> mutate(percent_loading = passengers/seats*100)

df13 |> 
  ggplot(aes(percent_loading)) +
  geom_histogram()+facet_wrap(~carrier_name, scales="free_y")
```

# Summarize the Data 


