---
title: "BOT Flight Data Analysis"
date: 09/22/2023
output: html_notebook
---

## Name: Alexis Alexander

This experiment (concerptually) continues our class work on flight data. 

We will continue to refer to (https://r4ds.hadley.nz/data-transform).

**However, we will expand and update our dataset by going to a root source -- the Bureau of Transportation.

Go to the website https://www.transtats.bts.gov/DL_SelectFields.aspx?gnoyr_VQ=FIM&QO_fu146_anzr=Nv4%20Pn44vr45. This dataset allows access to information on domestic flight and domestic carriers, with monthly summary data. 

# Packages


```{r}
library(nycflights13)
library(tidyverse)
library(janitor)
library(here)
```

# Get data from the online source.

Save the download to your data_raw directory. It will be a zipfile, but that will be OK.
Don't change the name on the zipfile.

## Load the BOT data 

We take advantage of the here package to get the path to the file that we stored. We also may not that inside the zipfile is a csv file. We will read the csv file. We unpackage all of the data and organized them in a dataframe based on the year 

```{r}

thisfile=here("data_raw", "DL_SelectFields_2017.zip")
df2017=read_csv(thisfile) %>% clean_names()

thisfile=here("data_raw", "DL_SelectFields_2018.zip")
df2018=read_csv(thisfile) %>% clean_names()

thisfile=here("data_raw", "DL_SelectFields_2019.zip")
df2019=read_csv(thisfile) %>% clean_names()

thisfile=here("data_raw", "DL_SelectFields_2020.zip")
df2020=read_csv(thisfile) %>% clean_names()

thisfile=here("data_raw", "DL_SelectFields_2021.zip")
df2021=read_csv(thisfile) %>% clean_names()

thisfile=here("data_raw", "DL_SelectFields_2022.zip")
df2022=read_csv(thisfile) %>% clean_names()

thisfile=here("data_raw", "DL_SelectFields_2023.zip")
df2023=read_csv(thisfile) %>% clean_names()
```

# Combine the Dataframes

Combining all of the rows into a single table from each data frame 

```{r}
dfc=df2017 |>
  bind_rows(df2018) |>
  bind_rows(df2019) |>
  bind_rows(df2020) |>
  bind_rows(df2021) |>
  bind_rows(df2022) |>
  bind_rows(df2023) 

```

### Create a new file 

I will save this data as csv file with intent that future loading cna simply read this new file. 
```{r}
write_csv(dfc,here("data","bot_flight_2017_2023m08.csv"))
```


# Create a Visualization 

For Arrivals and Depatures from LGA 

```{r}
dfcl=dfc |> 
  filter(passengers>0, origin=="LGA" | dest =="LGA")

```

Creating a graph for the total passengers by month over a year
```{r}

dfcl|> 
  ggplot(aes(month)) + 
  geom_bar(aes(weight=passengers)) +
  ggtitle("Total Passengers by Month over Multiyear Period")

```
 
 ## Year over Year7

```{r}
dfcl|> 
  ggplot(aes(year)) + 
  geom_bar(aes(weight=passengers)) +
  ggtitle("Total Passengers by Month over Multiyear Period")
```

# Task Give the month over month levels for the entire period from 2017-2023

```{r}
dfcl|> 
mutate(month_year = year+month/12) |>
  ggplot(aes(month_year)) +
    geom_bar(aes(weight=passengers))+
    ggtitle("Total Passengers by Month over Multiyear Period")
```

### Question - do more people travel "LGA to MIA" or "MIA to LGA"?
(Do we have a net migration?)

df2 is flights from LGA to MIA and vice versa 
dfcl is flights coming from or to LGA and then df2 was further filtered to find flights in LGA coming to and from Miami 

```{r}
df2=dfcl |> 
  filter(passengers>0, origin=="MIA" | dest =="MIA")
```

Month by month picture of people going from LGA to MIA and vice versa 

```{r}
df2|> 
mutate(month_year = year+month/12) |>
  ggplot(aes(month_year))+
  geom_bar(aes(weight=passengers, color=origin), position = position_dodge()) +
    ggtitle("Total Passengers by Month over Multiyear Period, between LGA and MIA")
```
## Split into passengers to MIA and passengers to LGA 

look up "grouping"
the difference between the flows and net migration, look at month by month as well as a total value 

Grouped by the origin and month and then summarize by the sum of passengers 

```{r}
  df2 %>%
  group_by(origin, month) %>%
  summarise(total_passengers=sum(passengers))
  mutate(cum_pass=sum(passengers))
  pivot_wider(names_from = origin, values_from = c("passengers","cum_pass")) %>%
  mutate(net_to_miami=passengers_LGA-passengers_MIA, cum_to_miami=cum_pass_LGA-cum_pass_MIA)
    
```

```{r}
df2 %>%
  group_by(origin, month) %>%
  summarise(total_passengers=sum(passengers))
  pivot_wider(names_from = origin, values_from = "total_passengers") %>%
  mutate(net_to_miami=passenger_LGA-passengers_MIA,cum_to_miami=cum_passenger_LGA-cum_passenger_MIA)
  ggplot()
```


## Notes Section From Class 
```{r}
df2|> 
mutate(month_year = year+month/12) |>
  ggplot(aes(month_year))+
  geom_line(aes(y=total_passengers), color=origin), position = position_dodge()) +
    ggtitle("Total Passengers by Month over Multiyear Period, between LGA and MIA")

mutate(cum_pass=cumsum(passengers)) %>%
  pivot_wider(names_from = origin,
              values_from = c("Passengers", "cum_pass")) %>%
  mutate(net_to_miami-passenger_LGA-passengers_MIA,)

df_f=dfcl%>%
  filter(dest =="MIA" | origin == "MIA") %>%
  group_by(origin,month) %>%
  summarise(passengers-sum(passengers))%>%
  mutate(cum_pass-cumsum(passengers)) %>%
  pivot_wider(names_from = origin,
              values_from = c("passengers", "cum_pass")) %>%
  mutate(net_to_miami=passenger_LGA-passengers_MIA, cum_to_miami=cum_passenger_LGA-cum_passenger_MIA)

```
Revise the analysis for Monday - grouping and filtering by states - All of new york to all of florida 

## Filtering the Data State-wise Instead of by Airport

df4 is the set of data representing all flights going to and from New York State

```{r}
df4=dfc |> 
  filter(passengers>0, origin_state_abr=="NY" | dest_state_abr =="NY")

```

df5 represents flights going from New York to Florida and vice versa 

```{r}
df5 =df4 |>
  filter(passengers>0, origin_state_abr=="FL" | dest_state_abr== "FL")
view(df5)

```


Month by month picture of people going from New York to Florida and vice versa 

```{r}
df5|> 
mutate(month_year = year+month/12) |>
  ggplot(aes(month_year))+
  geom_bar(aes(weight=passengers, color=origin_state_abr), position = position_dodge()) +
    ggtitle("Total Passengers by Month over Multiyear Period, between New York and Florida")
```

Grouped by the origin and month and then summarize by the sum of passengers 

```{r}
df5 %>%
  group_by(origin_state_abr, month) %>%
  summarise(total_passengers=sum(passengers))
```


