---
title: "EXPERIMENT 4_bikeshare data "
subtitle: "Effect of rain in ridership"
author: "Alexis Alexander"
date: "2023-11-03"
output: html_notebook
  toc: yes
---
This experiment seeks to understand how rain impacts the characteristics of riders. Potential area of investigation:

Does rain affect:
* How many people ride
* How long they ride
* How far they ride
* Where they ride

# Data Packages 

Baseline packages for our analysis

```{r}
library(tidyverse)
library(janitor)
library(here)
library(openmeteo)
```


# Loading the Bikeshare Raw Data
```{r}
df1 <- read_csv("data_raw/202309-capitalbikeshare-tripdata.csv", show_col_types = FALSE) %>% clean_names()
```

# Creating Dataframe 2 - Interjoing 
```{r}
df2s=df1 %>% 
  select(rideable_type,member_casual,
                	contains("start"),ride_id)%>% 
  mutate(start_stop="start")%>%
  rename(t=started_at,
     	station_name=start_station_name,
     	station_id=start_station_id,
     	lat=start_lat,
     	lng=start_lng)

df2e=df1 %>%
select(ride_id,rideable_type,member_casual,
                	contains("end")) %>%
  mutate(start_stop="stop") %>%
  rename(t=ended_at,
     	station_name=end_station_name,
     	station_id=end_station_id,
     	lat=end_lat,
     	lng=end_lng)

df2=bind_rows(df2s,df2e) %>%
  arrange(t) %>%
  mutate(rider_delta=(start_stop=="start")*2-1) %>% #change in ridership 
  mutate(riders=cumsum(rider_delta)) %>%
  relocate(riders,.after=t)

```

# Plotting the Dataframe

```{r}
df2 %>% 
  ggplot(aes(t,riders)) +
  geom_line()
```
# Creating a subsampled dataset 

```{r}
df_s=df2 %>% slice_head(n=1000)

```

Why looking at slicing to every one hundredth data point will be bad? - Will likely track features however will jump data points whether the data is sparse or not. Want the data to be equally spaced in time.

Starting with 100 rows to get the calculation right and then see if oyu can do it with a bigger data set. 
Data stored in number of seconds based on a reference point


Round down to the nearest 10 mins, relocate t_f next to t, and filter the data points by t_f

```{r}
df_e=df_s |>
  mutate(t_f=floor_date(t, "10 mins")) %>%
   relocate(t_f,.after=t) %>%
  slice_head(n=1,by=t_f)
```


# Applying to Previous Manipulations to the Entire Data Set 
```{r}
df_r=df2 |>
  mutate(t_f=floor_date(t,"10 mins")) %>%
  relocate(t_f,.after=t) %>%
  slice_head(n=1,by=t_f)
```

 
# Creating the Associated Plot 
```{r}

p1=df2 %>% 
  filter(day(t)==18) %>%
  ggplot(aes(x=t, y=riders)) +
  geom_line() +
  ggtitle("Riders on 18Sep")

p1+
  geom_line(data=df_r %>% filter(day(t)==18),
  color="red")

```
 

```{r}
df_r=df2 |>
  mutate(t_f=floor_date(t,"1 mins")) %>%
  relocate(t_f,.after=t) %>%
  slice_head(n=1,by=t_f)

```


```{r}
p1=df2 %>% 
  filter(day(t)==18) %>%
  ggplot(aes(x=t, y=riders)) +
  geom_line() +
  ggtitle("Riders on 18Sep")

p1+
  geom_line(data=df_r %>% filter(day(t)==18),
  color="red")
```

% Get Weather Data
```{r}
df_w=weather_history("Washington",
                	start = "2023-09-01",
                	end = "2023-09-30",
                	hourly = c("apparent_temperature",
                           	"wind_speed_10m",
                           	"precipitation")
)

```
# Merging bike and weather data

```{r}
df_s=df2 %>% slice_sample(n=1000)
df_j=df_s %>% left_join(df_w,
                     	by=join_by(closest(t>=datetime)))
```


```{r}
df_j=df_s %>% 
   left_join(df_w,by=join_by(closest(t>=datetime)))  %>%
   relocate(datetime, .after=t)
 
head(df_j)
```

## Investigating the Time Zone Mismatch Between the Bikeshare and Weather Datasets 

```{r}
df_j$t[1:5]
df_j$datetime[1:5]

```

```{r}
Sys.timezone()
```

```{r}
df_s_est <- force_tz(df_s, tzone = "America/New_York")
```


```{r}
df_j_est=df_s_est %>% 
   left_join(df_w,by=join_by(closest(t>=datetime)))  %>%
   relocate(datetime, .after=t)
 
head(df_j_est)
```

```{r}
df_j_est$t[1:5]
df_j_est$datetime[1:5]
```
```{r}
df2$t[1:5]
force_tz(df2$t[1:5],"America/New_York")
```

```{r}
df2c=df2 %>% mutate(t=force_tz(t,tzone="America/New_York")) #corrected
 
df_s2=df2c %>% slice_sample(n=1000)
 
df_j2=df_s2 %>% 
   left_join(df_w,by=join_by(closest(t>=datetime)))  %>%
   relocate(datetime, .after=t)
 
head(df_j2)

```

```{r}
dfc=df2c %>% 
  left_join(df_w,by=join_by(closest(t>=datetime))) %>%
   relocate(datetime, .after=t) %>%
  rename(atemp=hourly_apparent_temperature,
         wind=hourly_wind_speed_10m,
         prec=hourly_precipitation)
```


## A Visualization of the Data

Coloring when the rain is greater than 1- color each point based on whether it was raining or not 

```{r}
p2=dfc %>%
  ggplot(aes(x=t,y=riders,color=prec>1)) +
  geom_point()
p2
  
```


```{r}
p3 = dfc %>%
  filter(day(t)==23) %>%
  ggplot(aes(t,riders,color=wind,
             shape=rideable_type,
             size=member_casual))+
  geom_point() +
  ggtitle("Riders vs. Precipitation on Sep 23")

p3

```


```{r}
plotly::ggplotly(p3)
```

subtract the start and stop time to find the duration - between 2 date times
work with df1 with the initial data - 450000 rows with the start and end times 
call the variable ride_time with a histogram 

```{r}
df1_duration = df1 |>
  mutate(ride_time = ended_at-started_at)

df1_duration %>%
ggplot(aes(ride_time))+
  geom_histogram()

```

```{r}
df1_duration_c = df1_duration |>
  filter(ride_time >="60 sec" & ride_time <= "18000 sec")

```

Minimum and maximum cutoffs, bike became unusual during that time 
ignore anything greater than a 5 hour bike ride 
Filter away the extremely short and long rides and then redo the histogram 

```{r}

```





